ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-06-01 paper/Manuscript-Data-Package2.Rmd
Create a document with all manuscript changes and add answer letters for the reviewers.

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an `R` data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) of Canada's General Social Survey. Active travel is characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting."

@@

```{r rename-labels, include = FALSE}
# Renaming 'dest_label' and 'orig_label' values to reduce table space
gss_episodes_renamed <- gss_episodes %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

gss_impedances_renamed <- gss_impedances %>%
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))
```

@@

{ActiveCA} is distributed as an `R` package with a number of data 
objects and their documentation. `R` packages contain code, data, and documentation
in a standardized format that can be installed by `R` users via a software repository,
such as CRAN (Comprehensive R Archive Network) or GitHub, which makes them an 
adroit medium to distribute analysis-ready data.

@@

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29). Each row in the table correspond to a survey respondent, while the columns refer the following information: record identification (`PUMFID`), the person's weight (`WGHT_PER`), the month the survey data was collected (`SURVMNTH`), the respondent's age group (`AGEGR10`), the respondent's sex (`SEX`), and the respondent's marital status (`MARSTAT`).

The Main File of the 2015 GSS surveys includes a total of 17,390 respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and 848 variables.  For discrete variables, Statistics Canada has assigned specific codes to the possible values, with each code accompanied by a label. For instance, in the case of the variable `SURVMNTH`, a value of `1` means `January 2016`, 2 means `February 2016`, 3 corresponds to `March 2016`, and so on.

@@

```{r gss-main-file-2015}
main_2015_unprocessed <- gss_m_2015[1:10, 1:6]  %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.") %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `WGHT_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent."), threeparttable = T)

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an example of the Main File structure, Table \ref{tab:ep-2015-unprocessed} presents the first seven episodes for the record identification number `10041` and some variables from the GSS PUMF 2015 Episode File (Cycle 29). Each row in the table corresponds to an episode associated with the specified record identification (`PUMFID = 10041`), including the episode's weight (`WGHT_EPI`), episode number (`EPINO`), activity code (`TUI_01`), episode duration (`DURATION`), and episode location (`LOCATION`).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this respondent began the diary description by sleeping at home (`TUI_01 = 1` and `LOCATION = 300`) for 210 minutes, followed by 40 minutes of personal hygiene (`TUI_01 = 2`). The respondent then spent 15 minutes on personal care activities, such as getting ready for school, supervising homework, reading, playing, reprimanding, or providing educational or emotional support, as indicated by `TUI_01 = 27`. Next, they recorded a travel episode, walking for 15 minutes (`TUI_01 = 7` and `LOCATION = 315`), where both the origin and destination were their home. Such trips, where the journey starts and finishes at home, can be classified as recreational or leisure trips. Next, the respondent spent 3 hours searching for a job (`TUI_01 = 9`), took a 15-minute lunch break (`TUI_01 = 6`), and then cleaned the house (`TUI_01 = 18`) for two hours. Table \ref{tab:ep-2015-unprocessed} displays only six variables out of the 527 available. As shown, since the dataset does not label the variable values, decoding them can be both time-consuming and challenging.

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) %>%
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number `10041`.") %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

By running `?ActiveCA` in `R`, you can access the complete package documentation. The documentation includes a description of the package, information about the authors and maintainers, and an overview of all pre-processed data sets included in the package.

Figure \ref{fig:walking_documentation} displays the documentation for the walking episodes from the 2015 GSS (`walking_2015.rda`). The documentation provides a brief description of the dataset, instructions on how to load the package, as well as details about the file format and descriptions of all variables (already labeled).

@@

The GSS surveys apply a probability sampling methodology, in which each episode or person selected in the sample represents several other episodes or persons not in the sample. The number of episodes and persons represented by a episode or person is determined by the weight or weighting factor. Because of this, every estimates of the number of episodes or persons need to be calculated applying the corresponding weighting factors.

For instance, if someone wants to calculate the percentage of respondents from the 2015 GSS survey with active travel episodes, it is necessary to account for the person's weights. In 2015, the weight variable is represented by `WGHT_PER.` The code chunk below demonstrates how to obtain the percentage of people with active travel episodes by age group. It uses the `dplyr`package [@dplyr2023] to manipulate the data.

The process begins by creating a dataset that sums the population by age group. Then, it joins the 2015 episodes with the 2015 Main File. Note that in both operations, the code sums the person weight variable (`WGHT_PER`) to obtain the correct population values. After this, a new dataset, `Active_percentage`, is created by merging both previous datasets. The percentage of active travel episodes by age group is calculated by dividing the total population by the population with active trip episodes, then multiplying by 100 and rounding with 2 decimal places.

The result shows that the age group with the highest share of active trips is those between 15 and 20 years old, with almost 37%, followed by those between 25 and 34 years old with around 33%. There is a significant drop in percentage for the following groups, with the percentage falling to between 15% and 18% for the remaining age groups.

```{r percentages-act, echo=TRUE, out.width='80%'}
# Load the dplyr library for data manipulation
library('dplyr')

@@

# Calculate the total population for each age group
Total_population <- gss_main_2015 %>% 
  group_by(AGEGR10) %>% 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes %>% 
      filter(YEAR == 2015) %>% 
  left_join(gss_main_2015, by = c("PUMFID"="PUMFID")) %>% 
  group_by(AGEGR10) %>% 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population %>% 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) %>%
  mutate(
    Percentage = round(Active_population/Total_population*100,2))

# Display the final result
Active_percentage
```

@@

df_ep_processed <- gss_episodes %>% 
  group_by(YEAR) %>%
  summarise(Survey = 'Episode', 
            Count = n(), 
            Weighted = sum(WGHT_EPI)) %>%
  rename(Year = YEAR) %>%
  dplyr::select(Survey, Year, Count, Weighted) %>% 
  arrange(-Year)

@@

table_df_processed <- df_processed %>%
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:processed-obs}Total number and weighted sum of records processed.") %>%
  kable_styling(full_width = FALSE, font_size = 10) %>%
  add_header_above(c(" " = 2, "Observations" = ncol(df_processed) - 2)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29), displayed in \ref{tab:main-2015-unprocessed} before our processing. Table \ref{tab:ep-2015-processed} presents the walking episodes for the record identification number `10041` from the GSS PUMF 2015 Episode File (Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only the unique active travel episode appear in Table \ref{tab:ep-2015-processed} since the records were filtered to select cases with walking or cycling episodes. For both cases, Tables \ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain labeled variables, facilitating the interpretation of the data.

```{r gss-processed-file-2015, echo=FALSE, cache=FALSE}
main_2015_processed <- gss_main_2015[1:10, 1:6]  %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.") %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `WGHT_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent."), threeparttable = T)

@@

ep_2015_unprocessed <- gss_episodes_renamed |>
  filter(PUMFID == 10041 & YEAR == 2015) %>%
  dplyr::select(PUMFID, WGHT_EPI, act_label, DURATION, orig_label, dest_label, MODE) %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    caption = "\\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number `10041` of the 2015 GSS survey.", 
       col.names = c("PUMFID", "WGHT_EPI","Activity","Duration","Origin", "Destination", "Mode")) %>% 
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

```{r table-01, echo=FALSE, cache=FALSE}
# Calculate min, mean, and max statistic (duration)
stats_data <- gss_episodes_renamed %>%
  group_by(YEAR, MODE) %>%
  dplyr::summarise(
    min = min(DURATION),
    mean = sum(DURATION * WGHT_EPI)/sum(WGHT_EPI),
    median =  Hmisc::wtd.quantile(DURATION, weights = WGHT_EPI, probs = 0.5),
    max = max(DURATION),
    std = sd(DURATION),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(min, mean, median, max, std), names_to = "Statistic", values_to = "Value") %>%
  pivot_wider(names_from = YEAR, values_from = Value)

# Combine counts and stats into one data frame
stats_data <- stats_data %>%
  mutate(MODE = factor(MODE, levels = c("Walking", "Cycling"))) %>%
  arrange(MODE, Statistic) %>%
  mutate(across(where(is.numeric), ~ round(.x, 0)))

# Rename the 'MODE' column to 'Mode' before generating the table
stats_data <- stats_data %>%
  rename(Mode = MODE) %>%
  dplyr::select(Mode, Statistic, everything()) %>% 
   mutate(Statistic = case_when(
    Statistic == "max" ~ "Maximum",
    Statistic == "min" ~ "Minimum",
    Statistic == "mean" ~ "Mean",
    Statistic == "median" ~ "Median",
    Statistic == "std" ~ "Standard deviation",
    TRUE ~ Statistic))

# Creating a latex table
styled_ch03_table_01 <- stats_data %>%
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:table-01}Descriptive statistics for episodes with active transport records",  
        align = c('l', 'l', rep('c', ncol(stats_data) - 2))) %>%
  kable_styling(full_width = FALSE, font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 2, "Year" = ncol(stats_data) - 2)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

Table \ref{tab:table-02} and \ref{tab:table-03} provide descriptive
statistics for the two modes of transportation, split by destination
categories, from 1986 to 1998 and from 2005 to 2015, respectively. In
Table \ref{tab:table-02}, one can observed that in 1986 and 1992,
walking trips destined for `home` had the highest medians. However, by
1998, the highest medians shifted to trips to `work or school`, a
transition that also occurred for cycling trips between 1992 and 1998.
Table \ref{tab:table-03} indicates that the median duration for walking trips to
`home` and `work or school` remained at 10 minutes.

```{r table-02, echo=FALSE, cache=FALSE, warning=FALSE, results="asis"}
# Filter for selected years and calculate statistics of trip duration
episodes_stats_before <- gss_episodes %>%
  filter(YEAR < 2005) %>%
  group_by(YEAR, dest_label, MODE) %>%
  dplyr::summarise(
    count = sum(WGHT_EPI),
    min = min(DURATION),
    median = Hmisc::wtd.quantile(DURATION, weights = WGHT_EPI, probs = 0.5),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages of trips by mode/year 
total_counts <- episodes_stats_before %>% 
  group_by(YEAR, MODE) %>% 
  summarise(total_count = sum(count), 
            .groups = "drop")

@@

# Pivot data for the table layout
wide_data <- episodes_stats_before %>%
  pivot_wider(
    id_cols = c(dest_label, MODE),
    names_from = YEAR,
    values_from = c(min, median, max, Percentage)
  )  %>% 
  arrange(MODE, dest_label, desc(Percentage_1998))

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an R data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) of Canada's General Social Survey. Active travel is characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting."

@@

```{r rename-labels, include = FALSE}
# Renaming 'dest_label' and 'orig_label' values to reduce table space
gss_episodes_renamed <- gss_episodes |>
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))

gss_impedances_renamed <- gss_impedances |>
  mutate(dest_label = case_when(
    dest_label == "Grocery store, other stores or mall" ~ "Grocery store",
    dest_label == "Restaurant, bar or club" ~ "Restaurant",
    dest_label == "Sports centre, field or arena" ~ "Sport area",
    dest_label == "Medical, dental or other health clinic" ~ "Health clinic",
    dest_label == "In the neighbourhood" ~ "Neighbourhood",
    dest_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ dest_label  # Keep all other values as they are
  ), 
  orig_label = case_when(
    orig_label == "Grocery store, other stores or mall" ~ "Grocery store",
    orig_label == "Restaurant, bar or club" ~ "Restaurant",
    orig_label == "Sports centre, field or arena" ~ "Sport area",
    orig_label == "Medical, dental or other health clinic" ~ "Health clinic",
    orig_label == "In the neighbourhood" ~ "Neighbourhood",
    orig_label == "Library, museum or theatre" ~ "Cultural venues",
    TRUE ~ orig_label))
```

@@

{ActiveCA} is distributed as an R package with a number of data 
objects and their documentation. R packages contain code, data, and documentation
in a standardized format that can be installed by R users via a software repository,
such as CRAN (Comprehensive R Archive Network) or GitHub, which makes them an 
adroit medium to distribute analysis-ready data.

@@

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29). Each row in the table correspond to a survey respondent, while the columns refer the following information: record identification (PUMFID), the person's weight (WGHT_PER), the month the survey data was collected (SURVMNTH), the respondent's age group (AGEGR10), the respondent's sex (SEX), and the respondent's marital status (MARSTAT).

The Main File of the 2015 GSS surveys includes a total of 17,390 respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and 848 variables.  For discrete variables, Statistics Canada has assigned specific codes to the possible values, with each code accompanied by a label. For instance, in the case of the variable SURVMNTH, a value of 1 means January 2016, 2 means February 2016, 3 corresponds to March 2016, and so on.

@@

```{r gss-main-file-2015}
main_2015_unprocessed <- gss_m_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: `PUMFID`: record identification. `WGHT_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent."), threeparttable = T)

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an example of the Main File structure, Table \ref{tab:ep-2015-unprocessed} presents the first seven episodes for the record identification number 10041 and some variables from the GSS PUMF 2015 Episode File (Cycle 29). Each row in the table corresponds to an episode associated with the specified record identification (PUMFID = 10041), including the episode's weight (WGHT_EPI), episode number (EPINO), activity code (TUI_01), episode duration (DURATION), and episode location (LOCATION).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this respondent began the diary description by sleeping at home (TUI_01 = 1 and LOCATION = 300) for 210 minutes, followed by 40 minutes of personal hygiene (TUI_01 = 2). The respondent then spent 15 minutes on personal care activities, such as getting ready for school, supervising homework, reading, playing, reprimanding, or providing educational or emotional support, as indicated by TUI_01 = 27. Next, they recorded a travel episode, walking for 15 minutes (TUI_01 = 7 and LOCATION = 315), where both the origin and destination were their home. Such trips, where the journey starts and finishes at home, can be classified as recreational or leisure trips. Next, the respondent spent 3 hours searching for a job (TUI_01 = 9), took a 15-minute lunch break (TUI_01 = 6), and then cleaned the house (TUI_01 = 18) for two hours. Table \ref{tab:ep-2015-unprocessed} displays only six variables out of the 527 available. As shown, since the dataset does not label the variable values, decoding them can be both time-consuming and challenging.

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) |>
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number 10041.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

The documentation of the package can be accessed by running ?ActiveCA in R. The documentation includes a description of the package, information about the authors, contributors, and maintainers, and an overview of all pre-processed data sets included in the package.

Figure \ref{fig:walking_documentation} displays the documentation for the walking episodes from the 2015 GSS (walking_2015.rda). The documentation provides a brief description of the dataset, instructions on how to load the package, as well as details about the file format and descriptions of all variables (already labeled).

@@

The GSS surveys apply a probability sampling methodology, in which each episode or person selected in the sample represents several other episodes or persons not in the sample. The number of episodes and persons represented by a episode or person is determined by the weight or weighting factor. Because of this, estimates of the number of episodes or persons need to be calculated applying the corresponding weighting factors.

For instance, to calculate the percentage of respondents from the 2015 GSS survey with active travel episodes, it is necessary to account for the person's weight. In 2015, the weight variable is represented by WGHT_PER. The code below demonstrates how to obtain the percentage of people with active travel episodes by age group. It uses the dplyr package [@dplyr2023] to manipulate the data.

The process begins by creating a dataset that sums the population by age group. Then, it joins the 2015 episodes with the 2015 Main File. Note that in both operations, the code sums the person weight variable (WGHT_PER) to obtain the correct population values. After this, a new dataset, Active_percentage, is created by merging both previous datasets. The percentage of active travel episodes by age group is calculated by dividing the total population by the population with active trip episodes, then multiplying by 100 and rounding with 2 decimal places.

```
# Load the dplyr library for data manipulation
library('dplyr')

@@

# Calculate the total population for each age group
Total_population <- gss_main_2015 |> 
  group_by(AGEGR10) |> 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes |> 
      filter(YEAR == 2015) |> 
  left_join(gss_main_2015, by = c("PUMFID"="PUMFID")) |> 
  group_by(AGEGR10) |> 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population |> 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) |>
  mutate(
    Percentage = round(Active_population/Total_population*100,2))
```
The result shows that the age group with the highest share of active trips is those between 15 and 20 years old, with almost 37%, followed by those between 25 and 34 years old with around 33%. There is a significant drop in percentage for the following groups, with the percentage falling to between 15% and 18% for the remaining age groups.

```{r percentages-act, echo=FALSE, out.width='80%'}
# Load the dplyr library for data manipulation
library('dplyr')

# Calculate the total population for each age group
Total_population <- gss_main_2015 |> 
  group_by(AGEGR10) |> 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes |> 
      filter(YEAR == 2015) |> 
  left_join(gss_main_2015, by = c("PUMFID"="PUMFID")) |> 
  group_by(AGEGR10) |> 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population |> 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) |>
  mutate(
    Percentage = round(Active_population/Total_population*100,2))

# Display the final result
Active_percentage |>
  kable()
```

@@

df_ep_processed <- gss_episodes |> 
  group_by(YEAR) |>
  summarise(Survey = 'Episode', 
            Count = n(), 
            Weighted = sum(WGHT_EPI)) |>
  rename(Year = YEAR) |>
  dplyr::select(Survey, Year, Count, Weighted) |> 
  arrange(-Year)

@@

table_df_processed <- df_processed |>
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:processed-obs}Total number and weighted sum of records processed.") |>
  kable_styling(full_width = FALSE, font_size = 10) |>
  add_header_above(c(" " = 2, "Observations" = ncol(df_processed) - 2)) |>
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29), displayed in \ref{tab:main-2015-unprocessed} before our processing. Table \ref{tab:ep-2015-processed} presents the walking episodes for the record identification number 10041 from the GSS PUMF 2015 Episode File (Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only the unique active travel episode appear in Table \ref{tab:ep-2015-processed} since the records were filtered to select cases with walking or cycling episodes. For both cases, Tables \ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain labeled variables, facilitating the interpretation of the data.

```{r gss-processed-file-2015, echo=FALSE, cache=FALSE}
main_2015_processed <- gss_main_2015[1:10, 1:6]  |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: `PUMFID`: record identification. `WGHT_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent."), threeparttable = T)

@@

ep_2015_unprocessed <- gss_episodes_renamed |>
  filter(PUMFID == 10041 & YEAR == 2015) |>
  dplyr::select(PUMFID, WGHT_EPI, act_label, DURATION, orig_label, dest_label, MODE) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    caption = "\\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number `10041` of the 2015 GSS survey.", 
       col.names = c("PUMFID", "WGHT_EPI","Activity","Duration","Origin", "Destination", "Mode")) |> 
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

```{r table-01, echo=FALSE, cache=FALSE}
# Calculate min, mean, and max statistic (duration)
stats_data <- gss_episodes_renamed |>
  group_by(YEAR, MODE) |>
  dplyr::summarise(
    min = min(DURATION),
    mean = sum(DURATION * WGHT_EPI)/sum(WGHT_EPI),
    median =  Hmisc::wtd.quantile(DURATION, weights = WGHT_EPI, probs = 0.5),
    max = max(DURATION),
    std = sd(DURATION),
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(min, mean, median, max, std), names_to = "Statistic", values_to = "Value") |>
  pivot_wider(names_from = YEAR, values_from = Value)

# Combine counts and stats into one data frame
stats_data <- stats_data |>
  mutate(MODE = factor(MODE, levels = c("Walking", "Cycling"))) |>
  arrange(MODE, Statistic) |>
  mutate(across(where(is.numeric), ~ round(.x, 0)))

# Rename the 'MODE' column to 'Mode' before generating the table
stats_data <- stats_data |>
  rename(Mode = MODE) |>
  dplyr::select(Mode, Statistic, everything()) |> 
   mutate(Statistic = case_when(
    Statistic == "max" ~ "Maximum",
    Statistic == "min" ~ "Minimum",
    Statistic == "mean" ~ "Mean",
    Statistic == "median" ~ "Median",
    Statistic == "std" ~ "Standard deviation",
    TRUE ~ Statistic))

# Creating a latex table
styled_ch03_table_01 <- stats_data |>
  kable(format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "\\label{tab:table-01}Descriptive statistics for episodes with active transport records",  
        align = c('l', 'l', rep('c', ncol(stats_data) - 2))) |>
  kable_styling(full_width = FALSE, font_size = 10) |>
  column_spec(1, bold = TRUE) |>
  add_header_above(c(" " = 2, "Year" = ncol(stats_data) - 2)) |>
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")

@@

Table \ref{tab:table-02} and \ref{tab:table-03} provide descriptive
statistics for the two modes of transportation, split by destination
categories, from 1986 to 1998 and from 2005 to 2015, respectively. In
Table \ref{tab:table-02}, one can observed that in 1986 and 1992,
walking trips destined for home had the highest medians. However, by
1998, the highest medians shifted to trips to work or school, a
transition that also occurred for cycling trips between 1992 and 1998.
Table \ref{tab:table-03} indicates that the median duration for walking trips to
home and work or school remained at 10 minutes.

```{r table-02, echo=FALSE, cache=FALSE, warning=FALSE, results="asis"}
# Filter for selected years and calculate statistics of trip duration
episodes_stats_before <- gss_episodes |>
  filter(YEAR < 2005) |>
  group_by(YEAR, dest_label, MODE) |>
  dplyr::summarise(
    count = sum(WGHT_EPI),
    min = min(DURATION),
    median = Hmisc::wtd.quantile(DURATION, weights = WGHT_EPI, probs = 0.5),
    max = max(DURATION),
    .groups = "drop"
  )

# Calculate percentages of trips by mode/year 
total_counts <- episodes_stats_before |> 
  group_by(YEAR, MODE) |> 
  summarise(total_count = sum(count), 
            .groups = "drop")

@@

# Pivot data for the table layout
wide_data <- episodes_stats_before |>
  pivot_wider(
    id_cols = c(dest_label, MODE),
    names_from = YEAR,
    values_from = c(min, median, max, Percentage)
  )  |> 
  arrange(MODE, dest_label, desc(Percentage_1998))

wide_data <- wide_data |>
 dplyr::select(
    MODE, dest_label,
    min_1986, median_1986, max_1986, Percentage_1986,
    min_1992, median_1992, max_1992, Percentage_1992,
    min_1998, median_1998, max_1998, Percentage_1998
  ) |> 
  arrange(MODE, dest_label)
